---
title: "Pay_Gap"
author: "Raquel Baltazar"
date: "`r format(Sys.time(), '%a %d %b %Y (%H:%M:%S)')`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    theme: united
    highlight: tango
    toc: yes
    toc_depth: 3
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```
# Understanding the Data

## Gender pay gap in unadjusted form
% of average gross hourly earnings of men

The indicator measures the difference between average gross hourly earnings of male paid employees and of female paid employees as a percentage of average gross hourly earnings of male paid employees. The indicator has been defined as unadjusted, because it gives an overall picture of gender inequalities in terms of pay and measures a concept which is broader than the concept of equal pay for equal work. All employees working in firms with ten or more employees, without restrictions for age and hours worked, are included.

## Statistical Concept and Methodology

## Data Description 
The indicator measures the difference between average gross hourly earnings of male paid employees and of female paid employees as a percentage of average gross hourly earnings of male paid employees. The indicator has been defined as unadjusted, because it gives an overall picture of gender inequalities in terms of pay and measures a concept which is broader than the concept of equal pay for equal work. All employees working in firms with ten or more employees, without restrictions for age and hours worked, are included.

## Unit of Measure
% of average gross hourly earnings of men.

## Source 
ESS (SES)

Data source: Structure of Earnings Survey (SES)
Data provider: Eurostat, the statistical office of the European Union, based on data reported by the countries.
source link: https://ec.europa.eu/eurostat/databrowser/view/sdg_05_20/

**Copyrights:** Eurostat Copyright/Licence Policy is applicable.

## Other important information

online data code: SDG_05_20
Source of data: Eurostat
Last data update: 29/10/2020  22:00 (19 days ago)
Last structure update: 29/10/2020
Overall data coverage: 2002 — 2018
Number of values: 433


```{r library}

library(eurostat)
library(knitr)
library(kableExtra)
library(ggplot2)
library(magrittr)
library(data.table)
library(here)
library(countrycode)
library(gganimate)
theme_set(theme_bw())
library(gapminder)
library(data.table)

```

```{r query}
query <- search_eurostat(pattern = "Gender pay gap in unadjusted form", 
                         type = "table", fixed = FALSE)
query[, 1:2]
```


```{r new_variables}

ct <-  c("AT", "BE", "BG", "CH", "CY", "CZ", "DE", "DK", "EE", "EL", "ES", "FI", "FR", "HR","HU", "IE", "IS", "IT", "LI", "LT","LU", "LV", "MT", "NL", "NO", "PL", "PT","RO", "SE", "SI", "SK", "UK")

PTandEU <-  c("PT","EU27_2020")

```

```{r get}

dat <- get_eurostat(id="sdg_05_20", time_format = "num", 
                    filters = list (geo = ct))
dat [1:2,]

dat1<- get_eurostat(id="sdg_05_20", time_format = "num", 
                    filters = list (geo = PTandEU))
                    
dat1 [1:2,]
```


```{r label}
dat <- label_eurostat (dat)
dat[1:3,]


dat1 <- label_eurostat (dat1)
dat1[1:3,]
```


```{r clean_dat_dat1}

setDT(dat)

setDT(dat1)

dat[, c("unit", "nace_r2"):=NULL]

dat1[, c("unit", "nace_r2"):=NULL]

```

# Gender Pay Gap, 2003-2019

```{r eurostat_and_plots}
dat <- get_eurostat(id="sdg_05_20", filters = list(geo = ct))

library(ggplot2)
library(dplyr)
ggplot(dat, 
       aes(x = time, y= values, color = geo, label = geo)) + 
  geom_line (alpha = .5) +
  geom_text (data = dat %>% group_by(geo) %>% 
               filter(time == max(time)),
             size =2.6) +
  theme(legend.position = "none") + 
  labs(title = "Gender Pay Gap, 2003-2019",
       x= "Year", y= "%")

  labs(title = 'Year: {frame_time}', x = 'Time', y = 'Percentage') +
  transition_time(time) +
  ease_aes('linear')
```

# Animated

```{r animated_line}
ggplot(dat, aes(time, values, color = geo, label = geo)) +
  geom_line(alpha = .5) +
  geom_text (data = dat %>% group_by(geo) %>% 
               filter(time == max(time)),
             size =2.6) +
  theme(legend.position = "none") + 
  scale_color_viridis_d() +
  labs(x = "Year", y = "Gender Pay Gap") +
  theme(legend.position = "top") + 
  labs(title = "Gender Pay Gap",
       x= "Year", y= "%") +
  transition_reveal(time) +
  geom_point() +
  transition_reveal(time)

```

# Portugal vs. European Union, 2002-2019

```{r PT_vs_EU}

pteu <-  c("European Union - 27 countries (from 2020)","Portugal")

library(ggplot2)
library(dplyr)
ggplot(dat1, 
       aes(x = time, y= values, color = geo, label = geo)) + 
  geom_line (alpha = .5) +
  geom_text (data = dat1 %>% group_by(geo) %>% 
               filter(time == max(time)),
             size =2.6) +
  theme(legend.position = "none") + 
  labs(title = "Gender Pay Gap, 2002",
       x= "Year", y= "%")
```


# Then vs Now (bar chart and map)

## Gender Pay Gap in 2002

### Bar Chart

```{r bar_chart}
dat_2002 <- dat %>% 
  filter(time == "2002-01-01")
ggplot (dat_2002, aes(x= reorder(geo, values), y = values)) + 
  geom_col (color = "white", fill = "grey80") + 
  theme ( axis.text.x = element_text (size = 6)) + 
  labs (title = "Gender Pay Gap in 2002", 
        y = "%", x = NULL)
```

### Map
```{r map}

mapdata <-  get_eurostat_geospatial(nuts_level = 0) %>% 
  right_join (dat_2002) %>% 
  mutate (cat = cut_to_classes (values, n = 4, decimals = 1))
head(select(mapdata,geo,values,cat), 3)

```


```{r plot_map}

ggplot(mapdata, aes(fill = cat)) + 
  scale_fill_brewer(palette = "RdYlBu") + 
  geom_sf (color = alpha("white", 1/3), alpha = .6) + 
  xlim (c(-12,44)) + ylim(c(35, 70)) + 
  labs( title = "Gender Pay Gap in 2002",
        subtitle = "% of average gross hourly earnings of men",
        fill = "%")
```

**only a few countries have available data for 2002**
**Portugal only has available data from 2006 on**

## Let's try with 2006

### Bar chart
```{r bar_chart1}

library(gghighlight)

dat_2006 <- dat %>% 
  filter(time == "2006-01-01")
ggplot (dat_2006, aes(x= reorder(geo, values), y = values)) + 
  geom_col (color = "white", fill = "tomato") + 
  gghighlight(geo == "PT") +
  theme ( axis.text.x = element_text (size = 6)) + 
  labs (title = "Gender Pay Gap in 2006", 
        y = "%", x = NULL)
```

### Map
```{r map1}

mapdata <-  get_eurostat_geospatial(nuts_level = 0) %>% 
  right_join (dat_2006) %>% 
  mutate (cat = cut_to_classes (values, n = 4, decimals = 1))
head(select(mapdata,geo,values,cat), 3)

```


```{r plot_map1}

ggplot(mapdata, aes(fill = cat)) + 
  scale_fill_brewer(palette = "RdYlBu") + 
  geom_sf (color = alpha("white", 1/3), alpha = .6) + 
  xlim (c(-12,44)) + ylim(c(35, 70)) + 
  labs( title = "Gender Pay Gap in 2006",
        subtitle = "% of average gross hourly earnings of men",
        fill = "%")
```
## Gender Pay Gap in 2018

### Bar Chart

```{r bar_chart2}

dat_2018 <- dat %>% 
  filter(time == "2018-01-01")
ggplot (dat_2018, aes(x= reorder(geo, values), y = values)) + 
  geom_col (color = "white", fill = "tomato") + 
  gghighlight(geo == "PT") +
  theme ( axis.text.x = element_text (size = 6)) + 
  labs (title = "Gender Pay Gap in 2018", 
        y = "%", x = NULL)
```


### Map
```{r map2}

mapdata <-  get_eurostat_geospatial(nuts_level = 0) %>% 
  right_join (dat_2018) %>% 
  mutate (cat = cut_to_classes (values, n = 4, decimals = 1))
head(select(mapdata,geo,values,cat), 3)

```


```{r plot_map2}

ggplot(mapdata, aes(fill = cat)) + 
  scale_fill_brewer(palette = "RdYlBu") + 
  geom_sf (color = alpha("white", 1/3), alpha = .6) + 
  xlim (c(-12,44)) + ylim(c(35, 70)) + 
  labs( title = "Gender Pay Gap in 2019",
        subtitle = "% of average gross hourly earnings of men",
        fill = "%")
```

# ANIMATIONS FROM 2002 TO 2019

## Bar chart

```{r animated_bar_chart1}

ggplot (dat, aes(x= reorder(geo, values), y = values)) + 
  geom_col (color = "white", fill = "grey80") + 
  theme ( axis.text.x = element_text (size = 6)) + 
  labs( title = "Gender Pay Gap in 2019",
        subtitle = "% of average gross hourly earnings of men",
        fill = "%") + 
  labs(title = 'Year: {frame_time}') +
  transition_time(time) +
  ease_aes()
```

## The Map is unable to animate

This last animation is not working properly