---
title: "Gender Pay Gap in the EU"
author: "Raquel Baltazar"
date: "`r format(Sys.time(), '%a %d %b %Y (%H:%M:%S)')`"
output:
  html_document:
    theme: united
    highlight: tango
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

**We will use Eurostat’s indicator data for “Gender pay gap in unadjusted form” as a case study when working with the data.table R package and apply a few animations using gganimate. The goal is to explore the geographical and time trends for the gender pay gap in the EU and compare Portugal with some EU's countries**

We will start by learning how to understand the data and extract it from the Eurostat package within R. We will then process in preparation for some exploratory analysis.*

# Objective
The objective is to look at the geographical and time trends in the data. We will
answer the following questions:

+ What are the time trends for Portugal?
+ How does Portugal compare to other European countries?
+ Which countries have the highest pay gap percentage in Europe by year?

# Understanding the Data

## Gender pay gap in unadjusted form
% of average gross hourly earnings of men

The indicator measures the difference between average gross hourly earnings of male paid employees and of female paid employees as a percentage of average gross hourly earnings of male paid employees. The indicator has been defined as unadjusted, because it gives an overall picture of gender inequalities in terms of pay and measures a concept which is broader than the concept of equal pay for equal work. All employees working in firms with ten or more employees, without restrictions for age and hours worked, are included.

## Statistical Concept and Methodology

## Data Description 
The indicator measures the difference between average gross hourly earnings of male paid employees and of female paid employees as a percentage of average gross hourly earnings of male paid employees. The indicator has been defined as unadjusted, because it gives an overall picture of gender inequalities in terms of pay and measures a concept which is broader than the concept of equal pay for equal work. All employees working in firms with ten or more employees, without restrictions for age and hours worked, are included.

Overall data coverage: 2002 — 2018
Number of values: 433

## Unit of Measure
% of average gross hourly earnings of men.

## Source 
ESS (SES)
Source of data: Eurostat
Data source: Structure of Earnings Survey (SES)
Data provider: Eurostat, the statistical office of the European Union, based on data reported by the countries.
source link: https://ec.europa.eu/eurostat/databrowser/view/sdg_05_20/
online data code: SDG_05_20

**Copyrights:** Eurostat Copyright/Licence Policy is applicable.

# Loading Libraries

```{r library}

library(eurostat)
library(knitr)
library(kableExtra)
library(ggplot2)
library(magrittr)
library(data.table)
library(here)
library(countrycode)
library(gganimate)
library(gapminder)
library(data.table)

```

# Importing the Data

```{r query}
query <- search_eurostat(pattern = "Gender pay gap in unadjusted form", 
                         type = "table", fixed = FALSE)
query[, 1:2]
```


```{r new_variables}

ct <-  c("AT", "BE", "BG", "CH", "CY", "CZ", "DE", "DK", "EE", "EL", "ES", "FI", "FR", "HR","HU", "IE", "IS", "IT", "LI", "LT","LU", "LV", "MT", "NL", "NO", "PL", "PT","RO", "SE", "SI", "SK", "UK")

PTandEU <-  c("PT","EU27_2020")

```

# Variables dat and dat1

## dat = Every country in the EU separately
## dat1 = Portugal vs EU

```{r get}

dat <- get_eurostat(id="sdg_05_20", time_format = "num", 
                    filters = list (geo = ct))
dat [1:2,]

dat1<- get_eurostat(id="sdg_05_20", time_format = "num", 
                    filters = list (geo = PTandEU))
                    
dat1 [1:2,]
```


```{r label}
dat <- label_eurostat (dat)
dat[1:3,]


dat1 <- label_eurostat (dat1)
dat1[1:3,]
```
# Clean the data  (remove columns)

```{r clean_dat_dat1}

setDT(dat)

setDT(dat1)

dat[, c("unit", "nace_r2"):=NULL]

dat1[, c("unit", "nace_r2"):=NULL]

```

# Evolution of Gender Pay Gap in EU, 2002-2018

(there is no available data for 2019)

## Line graph

```{r eurostat_and_plots}
dat <- get_eurostat(id="sdg_05_20", filters = list(geo = ct))

library(ggplot2)
library(dplyr)
ggplot(dat, 
       aes(x = time, y= values, color = geo, label = geo)) + 
  geom_line (alpha = .5) +
  geom_text (data = dat %>% group_by(geo) %>% 
               filter(time == max(time)),
             size =2.6) +
  theme(legend.position = "none") + 
  labs(title = "Gender Pay Gap, 2003-2018",
       x= "Year", y= "%")

  labs(title = 'Year: {frame_time}', x = 'Time', y = 'Percentage') +
  transition_time(time) +
  ease_aes('linear')
```

## The same line graph, but animated

```{r animated_line}
ggplot(dat, aes(time, values, color = geo, label = geo)) +
  geom_line(alpha = .5) +
  geom_text (data = dat %>% group_by(geo) %>% 
               filter(time == max(time)),
             size =2.6) +
  theme(legend.position = "none") + 
  scale_color_viridis_d() +
  labs(x = "Year", y = "Gender Pay Gap") +
  theme(legend.position = "top") + 
  labs(title = "Gender Pay Gap, 2002-2018",
       x= "Year", y= "%") +
  transition_reveal(time) +
  geom_point() +
  transition_reveal(time)

```

## Bar chart 

```{r animated_bar_chart1}

ggplot (dat, aes(x= reorder(geo, values), y = values)) + 
  geom_col (color = "white", fill = "grey80") + 
  theme ( axis.text.x = element_text (size = 6)) + 
  labs( title = "Gender Pay Gap in 2019",
        subtitle = "% of average gross hourly earnings of men",
        fill = "%") + 
  labs(title = 'Year: {frame_time}') +
  transition_time(time) +
  ease_aes()
```

# Portugal vs. European Union, 2006-2018


**Portugal has no available data until 2006 and the EU only has available data from 2010 on**
(again, there is  no available data for 2019)

# line graph
```{r PT_vs_EU}

pteu <-  c("European Union - 27 countries (from 2020)","Portugal")

library(ggplot2)
library(dplyr)
ggplot(dat1, 
       aes(x = time, y= values, color = geo, label = geo)) + 
  geom_line (alpha = .5) +
  geom_text (data = dat1 %>% group_by(geo) %>% 
               filter(time == max(time)),
             size =2.6) +
  theme(legend.position = "none") + 
  labs(title = "Gender Pay Gap, 2006-2019",
       x= "Year", y= "%")
```

# Then vs Now 
Now we compare the first and the latest years [2002 vs. 2019] provided in the dataset individually.  

## Gender Pay Gap in 2002

### Bar Chart

```{r bar_chart}
dat_2002 <- dat %>% 
  filter(time == "2002-01-01")
ggplot (dat_2002, aes(x= reorder(geo, values), y = values)) + 
  geom_col (color = "white", fill = "grey80") + 
  theme ( axis.text.x = element_text (size = 6)) + 
  labs (title = "Gender Pay Gap in 2002", 
        y = "%", x = NULL)
```

### Map
```{r map}

mapdata <-  get_eurostat_geospatial(nuts_level = 0) %>% 
  right_join (dat_2002) %>% 
  mutate (cat = cut_to_classes (values, n = 4, decimals = 1))
head(select(mapdata,geo,values,cat), 3)

```


```{r plot_map}

ggplot(mapdata, aes(fill = cat)) + 
  scale_fill_brewer(palette = "RdYlBu") + 
  geom_sf (color = alpha("white", 1/3), alpha = .6) + 
  xlim (c(-12,44)) + ylim(c(35, 70)) + 
  labs( title = "Gender Pay Gap in 2002",
        subtitle = "% of average gross hourly earnings of men",
        fill = "%")
```


**only a few countries have available data for 2002**
**Portugal only has available data from 2006 on**

## Let's try using 2006 as a starting point

### Bar chart
```{r bar_chart1}

library(gghighlight)

dat_2006 <- dat %>% 
  filter(time == "2006-01-01")
ggplot (dat_2006, aes(x= reorder(geo, values), y = values)) + 
  geom_col (color = "white", fill = "tomato") + 
  gghighlight(geo == "PT") +
  theme ( axis.text.x = element_text (size = 6)) + 
  labs (title = "Gender Pay Gap in 2006", 
        y = "%", x = NULL)
```

**Portugal is highlighted for analysis and comparison**

### Map
```{r map1}

mapdata <-  get_eurostat_geospatial(nuts_level = 0) %>% 
  right_join (dat_2006) %>% 
  mutate (cat = cut_to_classes (values, n = 4, decimals = 1))
head(select(mapdata,geo,values,cat), 3)

```


```{r plot_map1}

ggplot(mapdata, aes(fill = cat)) + 
  scale_fill_brewer(palette = "RdYlBu") + 
  geom_sf (color = alpha("white", 1/3), alpha = .6) + 
  xlim (c(-12,44)) + ylim(c(35, 70)) + 
  labs( title = "Gender Pay Gap in 2006",
        subtitle = "% of average gross hourly earnings of men",
        fill = "%")
```
## Gender Pay Gap in 2018

(again, there is  no available data for 2019)

### Bar Chart

```{r bar_chart2}

dat_2018 <- dat %>% 
  filter(time == "2018-01-01")
ggplot (dat_2018, aes(x= reorder(geo, values), y = values)) + 
  geom_col (color = "white", fill = "tomato") + 
  gghighlight(geo == "PT") +
  theme ( axis.text.x = element_text (size = 6)) + 
  labs (title = "Gender Pay Gap in 2018", 
        y = "%", x = NULL)
```

### Map
```{r map2}

mapdata <-  get_eurostat_geospatial(nuts_level = 0) %>% 
  right_join (dat_2018) %>% 
  mutate (cat = cut_to_classes (values, n = 4, decimals = 1))
head(select(mapdata,geo,values,cat), 3)

```


```{r plot_map2}

ggplot(mapdata, aes(fill = cat)) + 
  scale_fill_brewer(palette = "RdYlBu") + 
  geom_sf (color = alpha("white", 1/3), alpha = .6) + 
  xlim (c(-12,44)) + ylim(c(35, 70)) + 
  labs( title = "Gender Pay Gap in 2018",
        subtitle = "% of average gross hourly earnings of men",
        fill = "%")
```


